Assignment 1: Data and Descriptive

### 2023-10-22 assignment_1.tex
Update 22.10

---
title: "Assignment 1: Data and Descriptive"
author: "Anine Therese Karlsen & Mona Lisa Jones"
format: pdf
abstract: "This assignment aims to acquire, process, and analyze sub-national GDP and population data for a subset of European countries. Calculate GDP per capita and explore regional inequity using various descriptive statistics and visualizations."
editor: visual
bibliography: [Econometrics.bib]
page-layout: full
number-sections: true
papersize: a4
---

```{r}
#| label: setup
#| output: false
#| echo: false

library(tidyverse)
library(tinytex)
library(ggplot2)
library(readxl)
library(magrittr)
library(vtable)
library(lubridate)
library(dineq)
```

# Introduction

While national GDP and GDP per capita are vital indicators of a country's aggregate economic health, they do not shed light on how wealth or income is distributed among its residents. A high national GDP can, paradoxically, coexist with pockets of regional deprivation [@lessmann2017].

The truth of this statement becomes more evident when taking a closer look in to into sub-national data. Regional wealth disparities are of prime concern, especially when crafting policies for equitable growth [@lessmann2017]. A country's macro-level prosperity does not automatically guarantee that all its regions partake equally in this wealth. By studying smaller regions within a country, it is possible to get a more nuanced narrative about the state of regional economic disparities [@lessmann2017].

@@

# Literature review

"In recent decades, the regional distribution of incomes within countries has attracted considerable interest among academics and policy makers." [@lessmann2017]

Slovakia, experienced significant economic growth. Their GDP per capita grew faster than the older EU member states, albeit from a lower starting point, leading to a convergence [@fullart].

The global financial crisis and subsequent European debt crisis had significant impacts on European economies. GDP per capita declined in several countries, particularly those hit hardest by the crisis, such as Greece, Spain, and Portugal. Some regions took many years to recover to pre-crisis levels, while others, especially in Northern Europe, recovered more quickly [@lewis].

Major metropolitan areas, like Paris in France, London in the UK, and Berlin in Germany, often grew faster than other regions in their respective countries, leading to increasing spatial disparities [@kiuru2019].

Denmark maintained relatively high levels of GDP per capita throughout this period, benefiting from a diversified economy, strong institutions, and a high degree of economic openness. It did face challenges during the 2008 economic crisis but recovered relatively quickly [@dynamics].

Slovenia, which joined the EU in 2004, showed growth in GDP per capita, though it too was affected by the global economic downturn in 2008. However, it managed to recover in the subsequent years [@fullart].

Regional disparities persisted, with the Île-de-France region (which includes Paris) significantly outperforming other regions in terms of GDP per capita. The gap between urban and rural areas also remained a topic of discussion [@regional].

Towards the end of this period, in early 2020, the COVID-19 pandemic posed a new set of challenges for European economies, causing significant contractions in GDP per capita across many regions [@regional].

# Part A: Sub-national GDP and GDP per Capita

## Data Acquisition and datasets

The first thing we did in this assignment was to acquire our data. Trough Eurostat, we could download the datasets nama_10r_3gdp and demo_r_pjanggr3 as csv files, as well as filter the data by our preferences before downloading it. As per the assignment, we filtered the dataset by choosing the years 2000 to 2020. Furthermore, we selected the NUTS 3 region for the nations we were given, which were Portugal, France, Hungary, Slovakia and Denmark. Finally, we specified the data to be in million Euro.

### GDP (nama_10r_3gdp)

The nama_10r_3gdp dataset from Eurostat provides insights into GDP at regional level using the NUTS classification system. It furnishes GDP values in both current prices and adjusted for inflation, with some figures given in purchasing power standards (PPS) to account for price level differences between countries. The data is often structured by year and region @eurostat2023.

@@

The different measures for the regional GDP are absolute figures in € and Purchasing Power Standards (PPS), figures per inhabitant and relative data compared to the EU Member States average @eurostat2023c.

### Population (demo_r_pjanggr3)

Using the NUTS categorization once more, Eurostat's demo_r_pjanggr3 records annual population changes at the regional level. This dataset includes information on births, deaths, net migration, and may also include demographic information on age and gender. It's also often displayed in a year-by-region format, and the data usually spans in yearly intervals @eurostat2023a.

@@

For the purpose of this research, the demographic data references the NUTS 2016 classification, which provides a detailed breakdown of the European Union's territory @eurostat2021.

### NUTS classification

The Nomenclature of Territorial Units for Statistics (NUTS) offers a stratified system to segment the economic territory of the EU (including the UK) to facilitate the consistent collection and harmonization of regional statistics across Europe. The NUTS regions range from NUTS 0 Country level to NUTS 3 small units such as municipalities level.

```{r}
#| echo: false
#| output: false

# This section of the code is dedicated to adding data to RStudio. We will import the datasets into the RStudio environment for further analysis, manipulation, and visualization using R.

# Data for population
pop_data <- read_csv("Raw_data/demo.gz")

# Data for GDP
gdp_data <- read_csv("Raw_data/nama.gz")
```

```{r}
#| echo: false
#| output: false
# Renaming and selecting the variables we want to keep
# We're doing this in order to make the datasets more clean and easy to read

gdp_data_01 <- gdp_data |>
  select(geo, year = TIME_PERIOD, GDP = OBS_VALUE)

pop_data_01 <- pop_data |>
  select(geo, year = TIME_PERIOD, population = OBS_VALUE)
```

## GDP per Capita Calculation

@@

$y_i=GDP_i/population_i$

```{r}
#| echo: false
#| output: false

# Merging the datasets pop_data and gdp_data into one dataset by joining the variables "geo" and "year".

merged_data <- left_join(pop_data_01, gdp_data_01, join_by(geo, year))

# Select Only NUTS 3 Regions
# Filter for specific countries, this will be useful when we will find the GDP per capita for the countries seperately

merged_data <- merged_data |>
  filter(
    startsWith(as.character(geo), "PT") | 
    startsWith(as.character(geo), "FR") | 
    startsWith(as.character(geo), "HU") |
    startsWith(as.character(geo), "SK") | 
    startsWith(as.character(geo), "DK")
  ) %>%

# Filter rows where 'geo' column has more than 4 characters to select only NUTS3 regions
  filter(nchar(as.character(geo)) > 4) %>%

# Rename 'geo' column to 'id_nuts3'
  rename(id_nuts3 = geo)
```

```{r}
#| echo: false
#| output: false

# Calculating GDP per Capita

GDP_per_capita <- merged_data |>
  mutate(
    GDP_per_capita = round(100*GDP/population, digits = 2)
  )

# we use *100 in order to get the gdp per capita as a percentage, and digits = 2 to limit it down to two decimals
```

@@

### Descriptive statistics

@@

"Interest in income inequality has led to the development of several ways of measuring it. Two types of measures are of interest in this paper----static and dynamic. *Static measures* provide a snapshot of these inequalities at a point of time whereas the dynamic *measures capture historical trends.*" (læreboken)

In this part, we'll look at GDP per capita for our assigned countries on a NUTS 3 level. In addition, we'll use different kinds of descriptive statistics in order to further 0,37, is analyse this data. In this analysis, we'll use @wooldridge2020 for help.

@@

**Mean**

@@

**MMR**

@@

**Standard diviation (SD)**

Calculating SD to quantify the dispersion or variability of a data set around its mean. Helping us assess the degree of uncertainty, variability, or risk associated with an economic variable or parameter, which is crucial for understanding the reliability of estimations and predictions [@wooldridge2020].

**Median**

The median serves as a robust measure of central tendency, especially when a dataset may have outliers or is skewed. Unlike the mean, the median is not influenced by extreme values and, thus, can provide a clearer picture of the "typical" value in situations where the data distribution is not symmetrical [@wooldridge2020].

### Portugal

```{r}
#| echo: false
#| output: false

# Making a dataset that only contains data from Portugal
# Use the na.omit function in order to remove NA values, cant make ggplot with NA values.

Portugal_data <- GDP_per_capita |>
  filter(grepl("PT", id_nuts3)) |>
  na.omit(Portugal_data)
```

```{r}
#| echo: false

# Making a figure using ggplot that shows us variations of GDP per capita across the different NUTS 3 regions in Portugal.

Portugal_data_plot <- ggplot(Portugal_data, aes(x = year, y = GDP_per_capita, color = id_nuts3)) +
  geom_line() +
  labs(title="Figure 1: GDP per Capita for Portugal", 
       x="Year",
       y="GDP per Capita") +
  theme(legend.position="none")

# Print out the figure

print(Portugal_data_plot)
```

```{r}
#| echo: false

# Calculate summary statistics for Portugal, chosing only GDP per capita 
summary_stats_portugal <- sapply(Portugal_data["GDP_per_capita"], function(x) c(
  mean = mean(x),
  median = median(x),
  std_dev = sd(x),
  minimum = min(x),
  maximum = max(x)
))

# Print the summary statistics
print(summary_stats_portugal)
```

@@

### France

```{r}
#| echo: false
#| output: false

# Making a dataset that only contains data from France

France_data <- GDP_per_capita |>
  filter(grepl("FR", id_nuts3)) |>
  na.omit(France_data)
```

```{r}
#| echo: false

# Making a figure using ggplot that shows us variations of GDP per capita across the different NUTS 3 regions in France.

France_data_plot <- ggplot(France_data, aes(x = year, y = GDP_per_capita, color = id_nuts3)) +
  geom_line() +
  labs(title="Figure 1: GDP per Capita for France", 
       x="Year",
       y="GDP per Capita") +
  theme(legend.position="none")

print(France_data_plot)
```

```{r}
#| echo: false

# Calculate summary statistics for France, chosing only GDP per capita 
summary_stats_france <- sapply(France_data["GDP_per_capita"], function(x) c(
  mean = mean(x),
  median = median(x),
  std_dev = sd(x),
  minimum = min(x),
  maximum = max(x)
))

# Print the summary statistics
print(summary_stats_france)
```

@@

### Hungary

```{r}
#| echo: false
#| output: false

# Making a dataset that only contains data from Hungary

Hungary_data <- GDP_per_capita |>
  filter(grepl("HU", id_nuts3)) |>
  na.omit(Hungary_data)
```

```{r}
#| echo: false

# Making a figure using ggplot that shows us variations of GDP per capita across the different NUTS 3 regions in Hungary.

Hungary_data_plot <- ggplot(Hungary_data, aes(x = year, y = GDP_per_capita, color = id_nuts3)) +
  geom_line() +
  labs(title="Figure 1: GDP per Capita for Hungary", 
       x="Year",
       y="GDP per Capita") +
  theme(legend.position="none")

print(Hungary_data_plot)
```

```{r}
#| echo: false

# Calculate summary statistics for Hungary, chosing only GDP per capita 
summary_stats_hungary <- sapply(Hungary_data["GDP_per_capita"], function(x) c(
  mean = mean(x),
  median = median(x),
  std_dev = sd(x),
  minimum = min(x),
  maximum = max(x)
))

# Print the summary statistics
print(summary_stats_hungary)
```

@@

### Slovakia

```{r}
#| echo: false
#| output: false

# Making a dataset that only contains data from Slovakia

Slovakia_data <- GDP_per_capita |>
  filter(grepl("SK", id_nuts3)) |>
  na.omit(Slovakia_data)
```

```{r}
#| echo: false

# Making a figure using ggplot that shows us variations of GDP per capita across the different NUTS 3 regions in Slovakia.

Slovakia_data_plot <- ggplot(Slovakia_data, aes(x = year, y = GDP_per_capita, color = id_nuts3)) +
  geom_line() +
  labs(title="Figure 1: GDP per Capita for Slovakia", 
       x="Year",
       y="GDP per Capita") +
  theme(legend.position="none")

print(Slovakia_data_plot)
```

```{r}
#| echo: false

# Calculate summary statistics for Slovakia, chosing only GDP per capita 
summary_stats_slovakia <- sapply(Slovakia_data["GDP_per_capita"], function(x) c(
  mean = mean(x),
  median = median(x),
  std_dev = sd(x),
  minimum = min(x),
  maximum = max(x)
))

# Print the summary statistics
print(summary_stats_slovakia)
```

@@

### Denmark

```{r}
#| echo: false
#| output: false

# Making a dataset that only contains data from Denmark

Denmark_data <- GDP_per_capita |>
  filter(grepl("DK", id_nuts3)) |>
  na.omit(Denmark_data)
```

```{r}
#| echo: false

# Making a figure using ggplot that shows us variations of GDP per capita across the different NUTS 3 regions in Denmark.

Denmark_data_plot <- ggplot(Denmark_data, aes(x = year, y = GDP_per_capita, color = id_nuts3)) +
  geom_line() +
  labs(title="Figure 1: GDP per Capita for Denmark", 
       x="Year",
       y="GDP per Capita") +
  theme(legend.position="none")

print(Denmark_data_plot)
```

```{r}
#| echo: false

# Calculate summary statistics for Denmark, chosing only GDP per capita 
summary_stats_denmark <- sapply(Denmark_data["GDP_per_capita"], function(x) c(
  mean = mean(x),
  median = median(x),
  std_dev = sd(x),
  minimum = min(x),
  maximum = max(x)
))

# Print the summary statistics
print(summary_stats_denmark)
```

@@

# Part B: Regional Inequity

## Gini Coefficient Calculation

@@

$GINW_j=\frac{1}{2 \bar{y_j}} \sum_{i}^{n_j}\sum_{l}^{n_j}\frac{p_i}{P_j} \frac{p_l}{P_j} |y_i-y_l|$

@@

The gini coefficient can help us measure inequality in a distribution, as is therefore a useful tool for us to use when we look at regional inequity. The closer the gini coefficient is to 1, the bigger the inequality is; a number closer to 0 equals equality. When looking at the gini coefficient for NUTS 2 regions, we also get a better overview over differences in income between different regions, and it also makes it easier to find the reasons as to why there is a difference between the regions @hasell2023.

```{r}
#| echo: false
#| output: false

# Before we can find the Gini coefficient for NUTS2 regions, we have to generate NUTS 2 and 0 identifiers

# Generate NUTS 2 and 0 Identifiers
main_data <- GDP_per_capita |> 
  mutate(
    id_nuts2 = substr(id_nuts3, 1, 4),
    id_nuts0 = substr(id_nuts3, 1, 2)
  ) 
```

```{r}
#| echo: false
#| output: false

# Calculating gini
gini <- main_data |>
  filter(population != 0) |>
  filter(population != "NA") |>
  filter(GDP_per_capita != 0) |>
  filter(GDP_per_capita != "NA") |>
  group_by(id_nuts2, year) |>
  mutate(gini_n2 = gini.wtd(GDP_per_capita, weights = population)) |>
  ungroup()

# Slice function to remove numbers that get repeated. 
gini_nuts_2 <- gini |>
  group_by(id_nuts2, year) |>
  slice(1) |>
  ungroup()

# Since some NUTS 2 and 3 regions are the same (like Copenhagen for example), some gini coefficients will be 0. We should remove them before doing further analysis and data presentation

# Filter data
gini_filtered <- gini_nuts_2 |>
  filter(gini_n2 != 0)
```

@@

```{r}
#| echo: false

# Creating a bar chart
ggplot(gini_filtered, aes(x = id_nuts2, y = gini_n2, fill = id_nuts0)) +
  geom_bar(stat = "identity") +
  ggtitle("Gini Coefficient by NUTS 2 Region") +
  xlab("NUTS 2 Region ID") +
  ylab("Gini Coefficient") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r}
#| echo: false

# Creating a time-series plot
ggplot(gini_filtered, aes(x = year, y = gini_n2, color = id_nuts2)) +
  geom_line() +
  ggtitle("Time-series Plot of Gini Coefficient by NUTS 2 Region") +
  xlab("Year") +
  ylab("Gini Coefficient")
```

```{r}
#| echo: false

# Creating a box plot
ggplot(gini_filtered, aes(x = id_nuts0, y = GDP_per_capita)) +
  geom_boxplot() +
  ggtitle("Distribution of GDP per Capita by Country") +
  xlab("Country") +
  ylab("GDP per Capita")
```

## Discussion

# References

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  a4paper,
  DIV=11,
  numbers=noendperiod]{scrartcl}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\KOMAoption{captions}{tableheading}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Assignment 1: Data and Descriptive},
  pdfauthor={Anine Therese Karlsen \& Mona Lisa Jones},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Assignment 1: Data and Descriptive}
\author{Anine Therese Karlsen \& Mona Lisa Jones}
\date{}

\begin{document}
\maketitle
\begin{abstract}
This assignment aims to acquire, process, and analyze sub-national GDP
and population data for a subset of European countries. Calculate GDP
per capita and explore regional inequity using various descriptive
statistics and visualizations.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, interior hidden, frame hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, enhanced, breakable]}{\end{tcolorbox}}\fi

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

While national GDP and GDP per capita are vital indicators of a
country's aggregate economic health, they do not shed light on how
wealth or income is distributed among its residents. A high national GDP
can, paradoxically, coexist with pockets of regional deprivation
(Lessmann and Seidel 2017).

The truth of this statement becomes more evident when taking a closer
look in to into sub-national data. Regional wealth disparities are of
prime concern, especially when crafting policies for equitable growth
(Lessmann and Seidel 2017). A country's macro-level prosperity does not
automatically guarantee that all its regions partake equally in this
wealth. By studying smaller regions within a country, it is possible to
get a more nuanced narrative about the state of regional economic
disparities (Lessmann and Seidel 2017).

@@

\hypertarget{literature-review}{%
\section{Literature review}\label{literature-review}}

``In recent decades, the regional distribution of incomes within
countries has attracted considerable interest among academics and policy
makers.'' (Lessmann and Seidel 2017)

Slovakia, experienced significant economic growth. Their GDP per capita
grew faster than the older EU member states, albeit from a lower
starting point, leading to a convergence ({``Full Article: The Causal
Impact of Economic Growth on Material Use in Europe,''} n.d.).

The global financial crisis and subsequent European debt crisis had
significant impacts on European economies. GDP per capita declined in
several countries, particularly those hit hardest by the crisis, such as
Greece, Spain, and Portugal. Some regions took many years to recover to
pre-crisis levels, while others, especially in Northern Europe,
recovered more quickly (Lewis and Verhoeven, n.d.).

Major metropolitan areas, like Paris in France, London in the UK, and
Berlin in Germany, often grew faster than other regions in their
respective countries, leading to increasing spatial disparities (Kiuru
and Inkinen 2019).

Denmark maintained relatively high levels of GDP per capita throughout
this period, benefiting from a diversified economy, strong institutions,
and a high degree of economic openness. It did face challenges during
the 2008 economic crisis but recovered relatively quickly ({``Dynamics
of Technological Innovation, Energy Consumption, Energy Price and
Economic Growth in Denmark - Murad - 2019 - Environmental Progress \&
Sustainable Energy - Wiley Online Library,''} n.d.).

Slovenia, which joined the EU in 2004, showed growth in GDP per capita,
though it too was affected by the global economic downturn in 2008.
However, it managed to recover in the subsequent years ({``Full Article:
The Causal Impact of Economic Growth on Material Use in Europe,''}
n.d.).

Regional disparities persisted, with the Île-de-France region (which
includes Paris) significantly outperforming other regions in terms of
GDP per capita. The gap between urban and rural areas also remained a
topic of discussion ({``Regional Inequality in Europe: Evidence, Theory
and Policy Implications \textbar{} Journal of Economic Geography
\textbar{} Oxford Academic,''} n.d.).

Towards the end of this period, in early 2020, the COVID-19 pandemic
posed a new set of challenges for European economies, causing
significant contractions in GDP per capita across many regions
({``Regional Inequality in Europe: Evidence, Theory and Policy
Implications \textbar{} Journal of Economic Geography \textbar{} Oxford
Academic,''} n.d.).

\hypertarget{part-a-sub-national-gdp-and-gdp-per-capita}{%
\section{Part A: Sub-national GDP and GDP per
Capita}\label{part-a-sub-national-gdp-and-gdp-per-capita}}

\hypertarget{data-acquisition-and-datasets}{%
\subsection{Data Acquisition and
datasets}\label{data-acquisition-and-datasets}}

The first thing we did in this assignment was to acquire our data.
Trough Eurostat, we could download the datasets nama\_10r\_3gdp and
demo\_r\_pjanggr3 as csv files, as well as filter the data by our
preferences before downloading it. As per the assignment, we filtered
the dataset by choosing the years 2000 to 2020. Furthermore, we selected
the NUTS 3 region for the nations we were given, which were Portugal,
France, Hungary, Slovakia and Denmark. Finally, we specified the data to
be in million Euro.

\hypertarget{gdp-nama_10r_3gdp}{%
\subsubsection{GDP (nama\_10r\_3gdp)}\label{gdp-nama_10r_3gdp}}

The nama\_10r\_3gdp dataset from Eurostat provides insights into GDP at
regional level using the NUTS classification system. It furnishes GDP
values in both current prices and adjusted for inflation, with some
figures given in purchasing power standards (PPS) to account for price
level differences between countries. The data is often structured by
year and region Eurostat (2023a).

@@

The different measures for the regional GDP are absolute figures in €
and Purchasing Power Standards (PPS), figures per inhabitant and
relative data compared to the EU Member States average Eurostat (2023b).

\hypertarget{population-demo_r_pjanggr3}{%
\subsubsection{Population
(demo\_r\_pjanggr3)}\label{population-demo_r_pjanggr3}}

Using the NUTS categorization once more, Eurostat's demo\_r\_pjanggr3
records annual population changes at the regional level. This dataset
includes information on births, deaths, net migration, and may also
include demographic information on age and gender. It's also often
displayed in a year-by-region format, and the data usually spans in
yearly intervals Eurostat (2023c).

@@

For the purpose of this research, the demographic data references the
NUTS 2016 classification, which provides a detailed breakdown of the
European Union's territory Eurostat (2021).

\hypertarget{nuts-classification}{%
\subsubsection{NUTS classification}\label{nuts-classification}}

The Nomenclature of Territorial Units for Statistics (NUTS) offers a
stratified system to segment the economic territory of the EU (including
the UK) to facilitate the consistent collection and harmonization of
regional statistics across Europe. The NUTS regions range from NUTS 0
Country level to NUTS 3 small units such as municipalities level. ~

\hypertarget{gdp-per-capita-calculation}{%
\subsection{GDP per Capita
Calculation}\label{gdp-per-capita-calculation}}

@@

\(y_i=GDP_i/population_i\)

@@

\hypertarget{descriptive-statistics}{%
\subsubsection{Descriptive statistics}\label{descriptive-statistics}}

@@

``Interest in income inequality has led to the development of several
ways of measuring it. Two types of measures are of interest in this
paper----static and dynamic. \emph{Static measures} provide a snapshot
of these inequalities at a point of time whereas the dynamic
\emph{measures capture historical trends.}'' (læreboken)

In this part, we'll look at GDP per capita for our assigned countries on
a NUTS 3 level. In addition, we'll use different kinds of descriptive
statistics in order to further 0,37, is analyse this data. In this
analysis, we'll use Wooldridge (2020) for help.

@@

\textbf{Mean}

@@

\textbf{MMR}

@@

\textbf{Standard diviation (SD)}

Calculating SD to quantify the dispersion or variability of a data set
around its mean. Helping us assess the degree of uncertainty,
variability, or risk associated with an economic variable or parameter,
which is crucial for understanding the reliability of estimations and
predictions (Wooldridge 2020).

\textbf{Median}

The median serves as a robust measure of central tendency, especially
when a dataset may have outliers or is skewed. Unlike the mean, the
median is not influenced by extreme values and, thus, can provide a
clearer picture of the ``typical'' value in situations where the data
distribution is not symmetrical (Wooldridge 2020).

\hypertarget{portugal}{%
\subsubsection{Portugal}\label{portugal}}

\includegraphics{assignment_1_files/figure-pdf/unnamed-chunk-7-1.pdf}

\begin{verbatim}
        GDP_per_capita
mean         1.4185524
median       1.3900000
std_dev      0.3702905
minimum      0.6500000
maximum      2.7200000
\end{verbatim}

@@

\hypertarget{france}{%
\subsubsection{France}\label{france}}

\includegraphics{assignment_1_files/figure-pdf/unnamed-chunk-10-1.pdf}

\begin{verbatim}
        GDP_per_capita
mean          2.630444
median        2.440000
std_dev       1.053767
minimum       0.830000
maximum      11.580000
\end{verbatim}

@@

\hypertarget{hungary}{%
\subsubsection{Hungary}\label{hungary}}

\includegraphics{assignment_1_files/figure-pdf/unnamed-chunk-13-1.pdf}

\begin{verbatim}
        GDP_per_capita
mean         0.8598000
median       0.7650000
std_dev      0.4059723
minimum      0.3200000
maximum      3.1300000
\end{verbatim}

@@

\hypertarget{slovakia}{%
\subsubsection{Slovakia}\label{slovakia}}

\includegraphics{assignment_1_files/figure-pdf/unnamed-chunk-16-1.pdf}

\begin{verbatim}
        GDP_per_capita
mean         1.2501948
median       1.0950000
std_dev      0.8018259
minimum      0.3000000
maximum      4.0200000
\end{verbatim}

@@

\hypertarget{denmark}{%
\subsubsection{Denmark}\label{denmark}}

\includegraphics{assignment_1_files/figure-pdf/unnamed-chunk-19-1.pdf}

\begin{verbatim}
        GDP_per_capita
mean          4.419221
median        4.000000
std_dev       1.343933
minimum       2.760000
maximum       8.420000
\end{verbatim}

@@

\hypertarget{part-b-regional-inequity}{%
\section{Part B: Regional Inequity}\label{part-b-regional-inequity}}

\hypertarget{gini-coefficient-calculation}{%
\subsection{Gini Coefficient
Calculation}\label{gini-coefficient-calculation}}

@@

\(GINW_j=\frac{1}{2 \bar{y_j}} \sum_{i}^{n_j}\sum_{l}^{n_j}\frac{p_i}{P_j} \frac{p_l}{P_j} |y_i-y_l|\)

@@

The gini coefficient can help us measure inequality in a distribution,
as is therefore a useful tool for us to use when we look at regional
inequity. The closer the gini coefficient is to 1, the bigger the
inequality is; a number closer to 0 equals equality. When looking at the
gini coefficient for NUTS 2 regions, we also get a better overview over
differences in income between different regions, and it also makes it
easier to find the reasons as to why there is a difference between the
regions Hasell and Roser (2023).

@@

\includegraphics{assignment_1_files/figure-pdf/unnamed-chunk-23-1.pdf}

\includegraphics{assignment_1_files/figure-pdf/unnamed-chunk-24-1.pdf}

\includegraphics{assignment_1_files/figure-pdf/unnamed-chunk-25-1.pdf}

\hypertarget{discussion}{%
\subsection{Discussion}\label{discussion}}

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-dynamics}{}}%
{``Dynamics of Technological Innovation, Energy Consumption, Energy
Price and Economic Growth in Denmark - Murad - 2019 - Environmental
Progress \& Sustainable Energy - Wiley Online Library.''} n.d.
\url{https://aiche.onlinelibrary.wiley.com/doi/full/10.1002/ep.12905}.

\leavevmode\vadjust pre{\hypertarget{ref-eurostat2021}{}}%
Eurostat. 2021. {``Population Change - {Demographic} Balance and Crude
Rates at Regional Level ({NUTS} 3)~ (Demo\_r\_gind3).''}
https://ec.europa.eu/eurostat/cache/metadata/en/demo\_r\_gind3\_esms.htm.

\leavevmode\vadjust pre{\hypertarget{ref-eurostat2023}{}}%
---------. 2023a. {``Gross Domestic Product ({GDP}) at Current Market
Prices by {NUTS} 3 Regions.''}
https://ec.europa.eu/eurostat/databrowser/view/nama\_10r\_3gdp/default/table?lang=en.

\leavevmode\vadjust pre{\hypertarget{ref-eurostat2023c}{}}%
---------. 2023b. {``Regional Economic Accounts (Reg\_eco10).''}
https://ec.europa.eu/eurostat/cache/metadata/en/reg\_eco10\_esms.htm.

\leavevmode\vadjust pre{\hypertarget{ref-eurostat2023a}{}}%
---------. 2023c. {``Population on 1 {January} by Broad Age Group, Sex
and {NUTS} 3 Region.''}
https://ec.europa.eu/eurostat/databrowser/view/demo\_r\_pjanaggr3/default/table?lang=en.

\leavevmode\vadjust pre{\hypertarget{ref-fullart}{}}%
{``Full Article: The Causal Impact of Economic Growth on Material Use in
Europe.''} n.d.
\url{https://www.tandfonline.com/doi/full/10.1080/21606544.2017.1325780}.

\leavevmode\vadjust pre{\hypertarget{ref-hasell2023}{}}%
Hasell, Joe, and Max Roser. 2023. {``Measuring Inequality: {What} Is the
{Gini} Coefficient?''} \emph{Our World in Data}, July.
